Stool Microbiome Analysis

### 2021-11-23 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Updates to workflow post meeting with collab. DESeq update and use original alpha div.

1) [Example using Negative Binomial in Microbiome Differential Abundance Testing](https://bioconductor.org/packages/devel/bioc/vignettes/phyloseq/inst/doc/phyloseq-mixture-models.html)
2) [Workflow for Microbiome Data Analysis: from raw reads to community analyses](https://bioconductor.org/help/course-materials/2017/BioC2017/Day1/Workshops/Microbiome/MicrobiomeWorkflowII.html#different_ordination_projections)

@@

```{r DESeq2 analysis}
# Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
sd <- sample_data(PSOtmao4G)
str(sd$sex)
str(sd$bmi_final.x)
# Factor sex, BMI okay
sample_data(PSOtmao4G)$sex = factor(sample_data(PSOtmao4G)$sex)

# DESeq analysis
mb <- phyloseq_to_deseq2(PSOtmao4G, ~ sex:age + bmi_final.x + tmao_mdn)
#mb2 <- estimateSizeFactors(mb, type="poscount")
mb3 = DESeq(mb, test="Wald", fitType="parametric")

# Results
res = results(mb3, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao4G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)
```

```{r Plot DESeq2 results}
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  ylab("Taxonomy") +
  xlab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

@@

We need to recalculate the alpha diversity metrics now that we have removed and glommed the taxa data. Note, by recalculating these metrics we don't want to use measures that count singletons (Chao1, ACE) because we purposefully removed those occurences! By redoing the metrics we are asking, "At the Genus level, is within diversity (richness, evenness, or a combination depending on the statistical test) different?"

```{r Update alpha statistics}
#plot_richness(PSOtmao4, x="tmao_mdn", color="bmi_cat", measures=c("Chao1", "Shannon"))

# Recalculate alpha diversity now that some taxonomy have been filtered out
updated_alpha <- estimate_richness(PSOtmao4G)
row.names(updated_alpha) <- gsub("X", "", row.names(updated_alpha))
```

First, load the master phenotype file so that we can have complete access to our covariate variables. Merge the master phenotyp file with the sample data from the phyloseq object.

```{r Alpha diversity load data}
# Load master pheno file
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")
phen <- subset(phen, select = -c(sex, age, age_cat, bmi_cat))

# Check bin structure and factor if needed
#str(phen$age_cat)
#phen$age_cat <- as.factor(phen$age_cat)
#str(phen$bmi_cat)phen$bmi_cat <- as.factor(phen$bmi_cat)

# Grab PSO sample data
df <- as(sample_data(PSOtmao4G), "data.frame")

# Merge
df2 <- merge(phen, df, by = 0)
row.names(df2) <- df2[,"Row.names"]
df2 <- subset(df2, select =-c(Row.names))

# Merge updated alpha
df_alpha2 <- merge(updated_alpha, df2, by = 0)
row.names(df_alpha2) <- df_alpha2[,"Row.names"]
df_alpha2 <- subset(df_alpha2, select =-c(Row.names))

# Original alpha
summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1, df2))
# Updated alpha
summary(lm(df_alpha2$Shannon ~ tmao_log + sex*age + cystatinc_bd1, df_alpha2))
```

Check the distributions of the alpha diversity metrics and run the multiple linear regression models.

```{r Original Alpha diversity stats}
# Check distribution
shapiro.test(df2$shannon) # not normal, proceed with caution
shapiro.test(log(df2$shannon))
# Run linear regression
summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$faith_pd) # normal
# Run linear regression
summary(lm(faith_pd ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$pielou_e) # not normal, proceed with caution
# Run linear regression
summary(lm(pielou_e ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$observed_otus) # normal
# Run linear regression
summary(lm(observed_otus ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
```

```{r Updated Alpha diversity stats}
#~~~~~~~~~~~
# Shannon
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Shannon) # not normal, proceed with caution
shapiro.test((df_alpha2$shannon)^.5) # not better
# Run linear regression
summary(lm(Shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# Fisher
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Fisher) # 
# Run linear regression
summary(lm(Fisher ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# InvSimpson
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$InvSimpson) # normal
# Run linear regression
summary(lm(InvSimpson ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# Observed
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Observed) # eh not normal
# Run linear regression
summary(lm(Observed ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
```

@@

```{r Original Alpha Diversity Plots}
# Shannon
y_variable <- "Shannon"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=shannon)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Faith's Phylogenetic Diversity
y_variable <- "Faith's Phylogenetic Diversity"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=faith_pd)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Pielou's Evenness
y_variable <- "Pielou's Evenness"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=pielou_e)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Observed OTUs
y_variable <- "Observed OTUs"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=observed_otus)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

```{r Updated Alpha Diversity Plots}
# Shannon
y_variable <- "Shannon"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Shannon)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Faith's Phylogenetic Diversity
y_variable <- "Chao1"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Chao1)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Pielou's Evenness
y_variable <- "Inverse Simpson"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=InvSimpson)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Observed OTUs
y_variable <- "Observed OTUs"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Observed)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

@@

# OTU table
MicroData <- otu_table(PSOtmao4G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao4G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df_alpha2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

@@

## ANCOM2 covariate analysis
ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?

```{r ANCOM - genus level with covariates}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao4G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao4G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df_alpha2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

@@

>>>>

1) [Example using Negative Binomial in Microbiome Differential Abundance Testing](https://bioconductor.org/packages/devel/bioc/vignettes/phyloseq/inst/doc/phyloseq-mixture-models.html)
2) [Workflow for Microbiome Data Analysis: from raw reads to community analyses](https://bioconductor.org/help/course-materials/2017/BioC2017/Day1/Workshops/Microbiome/MicrobiomeWorkflowII.html#different_ordination_projections)
3) [Example of what descriptive microbiome stats to report in the results](https://pubmed.ncbi.nlm.nih.gov/28407784/)

@@

From Dr. Kable's example,

> # Test the impact of fiber cluster, accounting for covariates, using the log ratio test
dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
>
> # This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero. 
dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
>
> # The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
resultsNames(dds.ratio.batcho)
res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))

## Factored by TMAO Median 
```{r DESeq2 analysis - TMAO median}
# Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
sd <- sample_data(PSOtmao1G)
str(sd$sex)
str(sd$bmi_final.x)
# Factor sex, BMI okay
sample_data(PSOtmao1G)$sex = factor(sample_data(PSOtmao1G)$sex)

#~~~~~~~~~~~~~~~
# DESeq analysis
# TMAO median
#~~~~~~~~~~~~~~~
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_mdn)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio)
alpha = 0.05
alpha.big = 0.5
sigtab = res.ratio[which(res.ratio$padj < alpha.big), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

No results when we look at TMAO median and control for age, sex, and BMI.

## Factored by TMAO Tertile

```{r DESeq2 analysis - TMAO tertile}

# DESeq analysis - tmao_tertile
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_tertile)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
alpha = 0.05
sigtab = res.ratio[which(res.ratio$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

#~~~~~~~~~~~~~~~
# Plot Results
#~~~~~~~~~~~~~~~
resOTU <- rownames(sigtab)
otuData <- as.data.frame(otu_table(PSOtmao1G))
otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
#otuData_desRes <- as.data.frame(t(otuData_desRes))
# Get sample data groupings
sampleData <- as.data.frame(sample_data(PSOtmao1G))
sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
# Get tax table
taxaData <- as.data.frame(tax_table(PSOtmao1G))
taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)

# Merge
ggData <- merge(taxaData_res, otuData_desRes, by = 0)
rownames(ggData) <-  ggData$OrderFamilyGenus
ggData <- ggData[,-(1:9)]
ggData <- as.data.frame(t(ggData))
ggData <- merge(ggData, sampleData, by = 0)
rownames(ggData) <- ggData[,"Row.names"]
ggData <- subset(ggData, select = -c(Row.names))

# Loop Plot
nres <- length(resOTU)
for (i in 1:nres) { 
    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
            geom_boxplot()+
            ylab(colnames(ggData)[i])) 
}

#~~~~~~~~~~~~~~~
# Plot Results Fold Change
#~~~~~~~~~~~~~~~
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  ylab("Taxonomy") +
  xlab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

## Factored by TMAO Quantile

```{r DESeq2 analysis - TMAO quantile}
# DESeq analysis - tmao_quantile
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
alpha = 0.05
sigtab = res.ratio[which(res.ratio$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

#~~~~~~~~~~~~~~~
# Plot Results
#~~~~~~~~~~~~~~~
resOTU <- rownames(sigtab)
otuData <- as.data.frame(otu_table(PSOtmao1G))
otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
#otuData_desRes <- as.data.frame(t(otuData_desRes))
# Get sample data groupings
sampleData <- as.data.frame(sample_data(PSOtmao1G))
sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
# Get tax table
taxaData <- as.data.frame(tax_table(PSOtmao1G))
taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)

# Merge
ggData <- merge(taxaData_res, otuData_desRes, by = 0)
rownames(ggData) <-  ggData$OrderFamilyGenus
ggData <- ggData[,-(1:9)]
ggData <- as.data.frame(t(ggData))
ggData <- merge(ggData, sampleData, by = 0)
rownames(ggData) <- ggData[,"Row.names"]
ggData <- subset(ggData, select = -c(Row.names))

# Loop Plot
nres <- length(resOTU)
for (i in 1:nres) { 
    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
            geom_boxplot()+
            ylab(colnames(ggData)[i])) 
}

#~~~~~~~~~~~~~~~
# Plot Results Fold Change
#~~~~~~~~~~~~~~~
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  ylab("Taxonomy") +
  xlab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

@@

We do not need to redo the alpha diversity calculations because the filtering we did was not to remove technical variation but instead to focus our analysis on the more common players (also so we can be more statistically confident using less scarce data). Note, if you did want to calculate alpha diversity, the command estimate_richness() can accomplish that for you.

Check the distributions of the alpha diversity metrics and run the multiple linear regression models with covariates.

```{r Original Alpha diversity stats}
# Load master pheno file
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")
phen <- subset(phen, select = -c(sex, age, age_cat, bmi_cat))

# Grab PSO sample data
df <- as(sample_data(PSOtmao1G), "data.frame")
# Merge
df2 <- merge(phen, df, by = 0)
row.names(df2) <- df2[,"Row.names"]
df2 <- subset(df2, select =-c(Row.names))

#~~~~~~~~~~~~~~~~
# Compute
#~~~~~~~~~~~~~~~~

# Check distribution
shapiro.test(df2$shannon) # not normal, proceed with caution
shapiro.test(log(df2$shannon))
# Run linear regression
summary(lm(tmao_log ~ shannon + sex*age + cystatinc_bd1, df2))

# Check distribution
shapiro.test(df2$faith_pd) # normal
# Run linear regression
summary(lm(tmao_log ~ faith_pd + sex*age + cystatinc_bd1, df2))

# Check distribution
shapiro.test(df2$pielou_e) # not normal, proceed with caution
# Run linear regression
summary(lm(tmao_log ~ pielou_e + sex*age + cystatinc_bd1, df2))

# Check distribution
shapiro.test(df2$observed_otus) # normal
# Run linear regression
summary(lm(tmao_log ~ observed_otus + sex*age + cystatinc_bd1, df2))
```

@@

```{r Original Alpha Diversity Plots}
# Shannon
y_variable <- "Shannon"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df2$tmao.x))
# Plot
ggplot(df2, aes(x=tmao_log, y=shannon)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Faith's Phylogenetic Diversity
y_variable <- "Faith's Phylogenetic Diversity"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df2$tmao.x))
# Plot
ggplot(df2, aes(x=tmao_log, y=faith_pd)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Pielou's Evenness
y_variable <- "Pielou's Evenness"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df2$tmao.x))
# Plot
ggplot(df2, aes(x=tmao_log, y=pielou_e)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Observed OTUs
y_variable <- "Observed OTUs"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df2$tmao.x))
# Plot
ggplot(df2, aes(x=tmao_log, y=observed_otus)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

@@

# OTU table
MicroData <- otu_table(PSOtmao1G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao1G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

@@

## ANCOM2 covariate analysis

ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?

```{r ANCOM - genus level with covariates median}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao1G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao1G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

@@

Repeat looking at tertile 1 and 3. This section isn't currently working but I'm not sure why.

```{r ANCOM - genus level with covariates tertile}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao1G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao1G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

# Step 2.1: Tertile 1 and 3
tert1_3 <- df2[df2$tmao_tertile == "Tertile1" | df2$tmao_tertile == "Tertile3",]
feature_table = feature_table[,colnames(feature_table) %in% rownames(tert1_3)]
meta_data = meta_data[rownames(meta_data) %in% rownames(tert1_3),]

# Define variables for ANCOM
main_var = "tmao_tertile"; p_adj_method = "BH"; alpha = 0.05
adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
# Run ANCCOM
t_start = Sys.time()
res = ANCOM(feature_table = feature_table, 
            meta_data = meta_data, 
            struc_zero = struc_zero, 
            main_var = main_var, 
            p_adj_method = p_adj_method, 
            alpha = alpha, 
            adj_formula = adj_formula, 
            rand_formula = rand_formula)
t_end = Sys.time()
t_run = t_end - t_start 
t_run # around 5s

# Merge with the phenotype data 
MetaData3 <- MetaData2 %>% dplyr::select(tmao_tertile)

# Gather the data
library(tidyr)
ggData2 <- gather(ggData,
                  key = "Taxa",
                  value = "Abundance",
                  -tmao_tertile)

# Plot it
p <- ggplot(data = ggData2,
            aes(x = Taxa, 
                y = Abundance,
                fill = tmao_tertile), 
            alpha = 0.1) +
  geom_boxplot(lwd=.25) +
  xlab(NULL) +
  ylab("Log of Taxa Abundance") +
  theme_bw() + # note order matters here - need bw to be before theme()
  theme(axis.text.y = element_text( hjust = 1),
        axis.text.x = element_text(angle = 60, hjust = 1),
        text=element_text(size=11))  +
  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
  #scale_fill_viridis(discrete = TRUE) +
  #coord_flip() +
  labs(fill = "TMAO \n Classification")
